\MakeUppercase{The Structure of $k$-Planar Graphs}

### 2019-07-05 k-planar.tex
First commit of David's version

\documentclass{patmorin}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{pat}
\usepackage{paralist}
\usepackage{stmaryrd}

\usepackage[longnamesfirst,numbers,sort&compress]{natbib}

@@

\DeclareMathOperator{\dist}{dist}
\DeclareMathOperator{\depth}{depth}
\DeclareMathOperator{\ff}{f}
\DeclareMathOperator{\tw}{tw}

\renewcommand{\proplabel}[1]{\label{prop:#1}}
\renewcommand{\propref}[1]{(PR\ref{prop:#1})}
\newcommand{\jlabel}[1]{\label{j:#1}}
\newcommand{\jref}[1]{(J\ref{j:#1})}
\newcommand{\tlabel}[1]{\label{t:#1}}
\newcommand{\tref}[1]{(T\ref{t:#1})}
\newcommand{\ylabel}[1]{\label{y:#1}}
\newcommand{\yref}[1]{(Y\ref{y:#1})}

@@

\title{\MakeUppercase{The Structure of $k$-Planar Graphs}%
    \thanks{This work was partly funded by NSERC and MRI.}}

\author{Vida Dujmović%
        \thanks{School of Computer Science and Electrical Engineering,
                University of Ottawa},\,\,
        Pat Morin%
        \thanks{School of Computer Science, Carleton University},\,\, and
        David R. Wood%
        \thanks{School of Mathematics, Monash University}}

%\renewcommand{\ge}{\geqslant}
%\renewcommand{\le}{\leqslant}

@@

\begin{abstract}
  Layered partitions of graphs (with small layered width, in which the quotient has small treewidth) are a recently-introduced tool that have been used to solve longstanding problems on queue layouts, non-repetitive colouring, and 3-d graph drawing.  Such partitions are known to exist for planar graphs and, more generally, apex-minor-free graphs.  In the current paper, we prove that every $k$-planar graph has a layered partition of layered width $O(k^2)$ in  which the quotient graph has treewidth $O(k^3)$. This is the first result of this type for a non-minor-closed class of graphs and implies that $k$-planar graphs have both queue number and non-repetitive chromatic number upper-bounded by a function of $k$. The former result was previously shown by combining layered partitions of planar graphs with \textit{ad hoc} methods. The latter result is new, for all $k\ge 1$.  These results extend to $(g,k)$-planar graphs, the natural generalization of $k$-planar graphs to to genus-$g$ surfaces (rather than the genus-$0$ plane).
\end{abstract}
\end{titlepage}

@@

In 1965, \citet{ringel:sechsfarbenproblem} introduced the notion of $1$-planar graphs, a generalization of planar graphs that allows edges to cross provided that no edge is involved in more than one crossing. This simple generalization has since become a rich source of problems and results.  Indeed, the annotated bibliography on 1-planar graphs by \cite{kobourov.liotta.ea:annotated} contains more than 140 entries.

The natural generalization of $1$-planar graphs, $k$-planar graphs (that allow edge to be involved in up to $k$ crossings), has been the topic of important recent work on graph drawing. Though planar graphs are, by now, quite well-understood, the situation is much less clear for $k$-planar graphs; even the maximum edge-density of $k$-planar graphs is not completely settled.  For $k\le 4$, the maximum number of edges in an $n$-vertex $k$-planar graph is $(k+3)(n-2)$, which is tight for $k=1,2$ \cite{pach.toth:graphs}.  A near-tight bound of $11(n-2)/2$ is known for $k=3$ \cite{pach.radoicic.ea:improving} but for $k>4$, the current best upper bound is $4.108\sqrt{k}n$ \cite{pach.toth:graphs}.

% In the current paper we investigate the structure of $k$-planar graphs.  Since planarity is closed under taking minors, planar graphs can be characterized by a finite set of forbidden minors.  Indeed, Wagner's Theorem states that this set consists of the complete 5-vertex graph $K_5$ and the complete bipartite graph $K_{3,3}$.  Unfortunately, $k$-planarity is not closed under taking minors, ruling out this kind of clean structural result.  The purpose of the current paper is to extend a new structural result for planar graphs to $k$-planar graphs.

The purpose of the current paper is to show that new structural results for planar graphs can be extended to $k$-planar graphs.  These results can be used to give a rough characterization of $k$-planar graphs of powers of bounded-degree planar graphs
% Rather, we present a rough structural characterization of $k$-planar graphs that, despite its roughness, provides several new results for $k$-planar and related graph families.  
All of these results are based on layered $H$-partitions, which we now define.
{\color{red}TODO: Help.}

A \emph{layering} of a graph $G$ is a sequence $\mathcal{L}=\langle V_0,V_1,\ldots\rangle$ such that $\{V_0,V_1,\ldots\}$ is a partition of $V(G)$ and there is no edge $vw\in E(G)$ such that $v\in V_i$, $w\in V_j$ and $|j-i|>1$.  For any partition $\mathcal{P}=\{S_1,\ldots,S_p\}$ of $V(G)$, a \emph{quotient graph} $H=G/\mathcal{P}$ has a $p$-element vertex set $V(H)=\{x_1,\ldots,x_p\}$ and $x_ix_j\in E(H)$ if and only if there exists an edge $vw\in E(G)$ such that $v\in S_i$ and $w\in S_j$. To highlight the importance of the quotient graph $H$, we call $\mathcal{P}$ an \emph{$H$-partition} and write this concisely as $\mathcal{P}=\{S_x : x\in V(H)\}$ so that each element of $\mathcal{P}$ is indexed by the vertex it creates in $H$.\footnote{An alternative definition of a layering of $G$ is an $H$-partition of $G$ where $H$ is a path.}  A layered $H$-partition $(\mathcal{L},\mathcal{P})$ of a graph $G$ consists of a layering $\mathcal{L}$ of $G$ and an $H$-partition $\mathcal{P}$ of $G$. The \emph{layered width} of $(\mathcal{L},\mathcal{P})$ is $\max\{|L\cap P|: L\in\mathcal{L},\, P\in\mathcal{P}\}$.

Layered $H$-partitions of small layered width in which $H$ has some additional property are a useful tool for a number of graph problems.  One such useful property is small \emph{treewidth}, which we now define. A \emph{tree-decomposition} $\mathcal{K}$ of a graph $G$ consists of a tree $K$ and a collection $\mathcal{K}=(B_x:x\in V(K))$ of subsets of $V(G)$ indexed by nodes of $K$ such that (i)~for every $vw\in E(G)$, there exists some $x\in V(K)$ with $v,w\in B_x$; and (ii)~for every $v\in V(G)$, the induced subgraph $K[x] := K[\{y: x\in B_y\}]$ is connected.  The \emph{width} of the tree-decomposition $\mathcal{K}$ is $\max\{|B_x|:x\in V(K)\}-1$.  The \emph{treewidth} $\tw(G)$ of a graph $G$ is the minimum width of a tree-decomposition of $G$.

 \citet{dujmovic.joret.ea:planar} show that, if each graph $G$ in some family $\mathcal{G}$ of graphs has a layered $H$-partition for which the layered width of the partition and the treewidth of $H$ are each upper-bounded by some constant $c_\mathcal{G}$, then each of the following quantities are upper bounded by a constant, for every $G\in\mathcal{G}$: queue number, track number, and layered treewidth. \citet{dujmovic.esperet.ea:planar} add non-repetitive chromatic number to this list.  \citet{dujmovic.joret.ea:planar} show that two common families of graphs have such $H$-partitions: planar graphs, graphs of bounded genus, and (more generally) apex-minor-free graphs.\footnote{A graph $M$ is a \textit{minor} of a graph $G$ if a graph isomorphic to $M$ can be obtained from a subgraph of $G$ by contracting edges. A class $\mathcal{G}$ of graphs is \emph{minor-closed} if for every graph $G\in\mathcal{G}$, every minor of $G$ is in $\mathcal{G}$. A minor-closed class is \emph{proper} if it is not the class of all graphs. For example, for fixed $g\geq 0$, the class of graphs with Euler genus at most $g$ is a proper minor-closed class.

 A graph $G$ is $t$-apex if it contains a set $A$ of at most $t$ vertices such that $G-A$ is planar. A 1-apex graph is \emph{apex}.  A minor-closed class $\mathcal{G}$ is apex-minor-free if some apex graph is not in $\mathcal{G}$.}

% In this paper we show that another family of graphs admits layered $H$-partitions of bounded layered width in which $H$ has bounded treewidth:
In the current paper we show that $k$-planar graphs also admit layered $H$-partitions in which $H$ has bounded treewidth. First, we formally define $k$-planar graphs.  An \emph{embedded graph} $G$ is a graph with $V(G)\subset\R^2$ in which each edge $vw\in E(G)$ is a closed curve\footnote{A closed curve in a surface $\Sigma$ is a continuous function $f:[0,1]\to \Sigma$. The points $f(0)$ and $f(1)$ are called the \emph{endpoints} of the curve.  When there is no danger of misunderstanding we treat a curve $f$ as the point set $\{f(t):0\le t\le 1\}$.} in $\R^2$ with endpoints $v$ and $w$ and not containing any vertex of $G$ in its interior.  A \emph{crossing} in an embedded graph $G$ is a triple $(p,vw,xy)$ with $p\in\R^2$, $vw,xy\in E(G)$ and such that $p\in (vw\cap xy)\setminus\{v,w,x,y\}$. An embedded graph $G$ is \emph{$k$-plane} if each edge of $G$ takes part in at most $k$ crossings.  A (not necessarily embedded) graph $G'$ is \emph{$k$-planar} if there exists $k$-plane graph $G$ isomorphic to $G'$.

Under these definitions, $0$-planar graphs are exactly planar graphs and $0$-plane graphs are exactly plane graphs. Thus, like genus-$g$ graphs,\footnote{The \textit{Euler genus} of the orientable surface with $h$ handles is $2h$. The \textit{Euler genus} of the non-orientable surface with $c$ cross-caps is $c$. The \textit{Euler genus} of a graph $G$ is the minimum integer $g$ such that $G$ embeds in a surface of Euler genus $g$. Of course, a graph is planar if and only if it has Euler genus 0; see \citep{mohar.thomassen:graphs} for more about graph embeddings in surfaces.} $k$-planar graphs are a generalization of planar graphs (which are genus-$0$ graphs).  Unlike genus-$g$ graphs, however, the family of $k$-planar graphs is not minor-closed.  A graph $G'$ obtained from a $k$-planar graph $G$ by edge deletions and edge contractions may or may not be $k$-planar. Indeed, \citet{dujmovic.eppstein.ea:structure} construct 1-planar graphs that contain arbitrarily large complete graph minors.

We prove the following result:
\begin{thm}\thmlabel{k-planar}
  Every $k$-planar graph has a layered $H$-partition of layered width at most $24k^2 + 60k + 36$ in which $H$ has treewidth at most $\treewidth$.
\end{thm}

@@

\begin{thm}\thmlabel{1-planar}
  Every 1-planar graph has a layered $H$-partition of layered width at most 30 where $H$ is planar and has treewidth at most 3.
\end{thm}

These results imply that $k$-planar graphs have all the graph parameters mentioned above upper-bounded by some constant $c_k$.

Prior to this work, the strongest structural characterization of $k$-planar graphs was in terms of layered treewidth, which we now define.  A \emph{layered tree-decomposition} $(\mathcal{L},\mathcal{K})$ consists of a layering $\mathcal{L}$ and a tree-decomposition $\mathcal{K}$ of $G$. The layered width of $(\mathcal{L},\mathcal{K})$ is $\max\{|L\cap B|: L\in \mathcal{L},\, B\in \mathcal{K}\}$.  The \emph{layered treewidth} of $G$ is the minimum layered width of any layered tree-decomposition of $G$.

>>>>

\documentclass{patmorin}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{amsfonts}
\usepackage{amsthm,mathtools}
\usepackage{pat}
\usepackage{paralist}
\usepackage{stmaryrd}

\usepackage[longnamesfirst,numbers,sort&compress]{natbib}
\usepackage[noabbrev,capitalise]{cleveref}
\allowdisplaybreaks
%\sloppy
%%%%%%%%%%
\makeatletter
\def\NAT@spacechar{~}
\makeatother
%%%%%%%%%%
\crefname{lem}{Lemma}{Lemmas}
\crefname{thm}{Theorem}{Theorems}
\crefname{cor}{Corollary}{Corollaries}
\crefname{conj}{Conjecture}{Conjectures}
\crefname{obs}{Observation}{Observations}
\crefname{prop}{Proposition}{Propositions}
\crefname{clm}{Claim}{Claims}
\crefname{openproblem}{Open Problem}{Open Problems}
\crefformat{equation}{(#2#1#3)}
\Crefformat{equation}{Equation #2(#1)#3}

@@

\DeclareMathOperator{\dist}{dist}
\DeclareMathOperator{\depth}{depth}
\DeclareMathOperator{\ff}{f}
\DeclareMathOperator{\tw}{tw}
\DeclareMathOperator{\pw}{pw}
\DeclareMathOperator{\qn}{qn}
\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}
\DeclarePairedDelimiter{\floor}{\lfloor}{\rfloor}

\newcommand{\PRlabel}[1]{\label{PR:#1}}
\newcommand{\PRref}[1]{(PR\ref{PR:#1})}
\newcommand{\jlabel}[1]{\label{j:#1}}
\newcommand{\jref}[1]{(J\ref{j:#1})}

\renewcommand{\ge}{\geqslant}
\renewcommand{\le}{\leqslant}
\renewcommand{\geq}{\geqslant}
\renewcommand{\leq}{\leqslant}

@@

\title{\MakeUppercase{The Structure of $k$-Planar Graphs}}

\author{Vida Dujmovi\'c%
        \thanks{School of Computer Science and Electrical Engineering,
                University of Ottawa, Ottawa, Canada (\texttt{vida.dujmovic@uottawa.ca}).
                Research  supported by NSERC and the Ontario Ministry of Research and Innovation.},\,\,
        Pat Morin%
        \thanks{School of Computer Science, Carleton University, Ottawa, Canada (\texttt{morin@scs.carleton.ca}).                 Research  supported by NSERC and the Ontario Ministry of Research and Innovation.},\,\, and
        David R. Wood\thanks{School of Mathematics, Monash University, Melbourne, Australia (\texttt{david.wood@monash.edu}). Research supported by the Australian Research Council.}
}

@@

\begin{abstract}
Dujmovi\'c~et~al.~(FOCS 2019) recently proved that every planar graph is a subgraph of the strong product of a graph of bounded treewidth and a path. This tool has been used to solve longstanding problems on queue layouts, non-repetitive colouring, and $p$-centred colouring. %, and graph encoding. 
We generalise this result for $k$-planar graphs, where a graph is \emph{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings. In particular, we prove that every $k$-planar graph is a subgraph of the strong product of a graph of treewidth $O(k^5)$ and a path. This is the first result of this type for a non-minor-closed class of graphs. It implies, amongst other results, that $k$-planar graphs have non-repetitive chromatic number upper-bounded by a function of $k$. All these results generalise for drawings of graphs on arbitrary surfaces. 
\end{abstract}
\end{titlepage}

% Layered partitions of graphs (with small layered width, in which the quotient has small treewidth) are a recently-introduced tool that have been used to solve longstanding problems on queue layouts, non-repetitive colouring, and 3-d graph drawing.  Such partitions are known to exist for planar graphs and, more generally, apex-minor-free graphs.  In the current paper, we prove that every $k$-planar graph has a layered partition of layered width $O(k^2)$ in  which the quotient graph has treewidth $O(k^3)$. This is the first result of this type for a non-minor-closed class of graphs and implies that $k$-planar graphs have both queue number and non-repetitive chromatic number upper-bounded by a function of $k$. The former result was previously shown by combining layered partitions of planar graphs with \textit{ad hoc} methods. The latter result is new, for all $k\ge 1$.  These results extend to $(g,k)$-planar graphs, the natural generalization of $k$-planar graphs to to genus-$g$ surfaces (rather than the genus-$0$ plane).

@@

A graph is \emph{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings. Such graphs provide a natural generalisation of planar graphs, and are important in graph drawing research; see the recent bibliography on 1-planar graphs and the 140 references therein \citep{kobourov.liotta.ea:annotated}. The present paper studies the structure of $k$-planar graphs (and other more general classes). Various structural results about planar graphs generalise for $k$-planar graphs. For example, \citet{FP08} generalised the Lipton-Tarjan separator theorem to show that every $k$-planar graph with $n$ vertices has a balanced separator of order $O(\sqrt{kn})$.

The main results of this paper generalise the following recent theorem of \citet{dujmovic.joret.ea:planar} to the setting of $k$-planar graphs (and other more general classes).

\begin{thm}[\citep{dujmovic.joret.ea:planar}] 
\label{PlanarBasic}
Every planar graph is a subgraph of $H\boxtimes P$, for some graph $H$ of treewidth at most $8$ and for some path $P$. 
\end{thm}

Here $\boxtimes$ is the strong product, and treewidth is an invariant that measures how `tree-like' a given graph is; see \cref{Tools} for formal definitions and see \cref{ProductExample} for an example.  So, loosely speaking, \cref{PlanarBasic} says that every planar graph is contained in the product of a tree-like graph and a path. This enables combinatorial results for graphs of bounded treewidth to be generalised for planar graphs (with different constants). \cref{kPlanarBasic} has been the key tool in solving several well-known open problems. In particular, \citet{dujmovic.joret.ea:planar} proved that planar graphs have bounded queue-number (resolving a conjecture of \citet{HLR92}), and \citet{dujmovic.esperet.ea:planar}  proved that planar graphs have bounded non-repetitive chromatic number (resolving a conjecture of \citet{AGHR-RSA02}).

\begin{figure}[!h]
\centering
\includegraphics{ProductExample}
\caption{Example of a strong product.}
\label{ProductExample}
\end{figure}

We generalise \cref{PlanarBasic} as follows.

\begin{thm}
\label{kPlanarBasic}
Every $k$-planar graph is a subgraph of $H\boxtimes P$, for some graph $H$ of treewidth at most $(4+o(1))k^5$ and for some path $P$. 
\end{thm}

This theorem and several related results are proved in \cref{Structure}. In particular, in \cref{sec-1-planar} we present a refinement of \cref{kPlanarBasic} in the case $k=1$. In \cref{sec-g-k-planar} we generalise \cref{kPlanarBasic} for graphs that have a drawing on an arbitrary surface with a bounded number of crossings per edge.

\cref{kPlanarBasic} (and its generalisations) have applications in diverse areas, including queue layouts  \citep{dujmovic.joret.ea:planar}, non-repetitive colouring  \citep{dujmovic.esperet.ea:planar}, and $p$-centred colouring  \citep{micek:personal}.
%, and graph encoding \citep{BGP}. 
These applications in the context of $k$-planar graphs are explored in \cref{Applications}. For example, we prove that $k$-planar graphs have bounded non-repetitive chromatic number (for fixed $k$). Prior to the recent work of 
\citet{dujmovic.esperet.ea:planar}, it was even open whether planar graphs have bounded non-repetitive chromatic number.

\cref{Characterisation} presents a rough characterisation of $k$-planar graphs in terms of an underlying planar graph. This is useful for showing that various families of graphs are $k$-planar. For example, we conclude that bounded powers of bounded-degree planar graphs are $k$-planar, for a suitable constant $k$. Thus 
\cref{kPlanarBasic} and the above applications are applicable to this class. These results also generalise for arbitrary surfaces.

\cref{Examples} presents several examples of $k$-planar graphs, including $k$-map graphs, $k$-string graphs, and $k$-nearest neighbour graphs. All of the above results are applicable to these examples.

%\citet{dujmovic.joret.ea:planar} generalised their result for planar graphs for graphs embeddable on any surface. In particular, they proved that every graph of Euler genus $g$ is a subgraph of $H\boxtimes P$ for some graph $H$ of treewidth at most $???k$ and for some path $P$. We provide a further generalisation allowing crossings. Say a graph is $(g,k)$-planar if it has a drawing in a surface of Euler genus at most $g$ in which each edge is involved in at most $k$ crossings. 
%
%\begin{thm}
%For all integers $g,k\geq 1$ there is an integer $c_{k,g}$, such that every $(g,k)$-planar graph is a subgraph of $H\boxtimes P$, for some graph $H$ of treewidth at most $c_{k,g}$  and for some path $P$. 
%\end{thm}

%In 1965, \citet{ringel:sechsfarbenproblem} introduced the notion of $1$-planar graphs, a generalization of planar graphs that allows edges to cross provided that no edge is involved in more than one crossing. This simple generalization has since become a rich source of problems and results.  Indeed, the annotated bibliography on 1-planar graphs by \cite{kobourov.liotta.ea:annotated} contains more than 140 entries.
%
%The natural generalization of $1$-planar graphs, $k$-planar graphs (that allow edge to be involved in up to $k$ crossings), has been the topic of important recent work on graph drawing. Though planar graphs are, by now, quite well-understood, the situation is much less clear for $k$-planar graphs; even the maximum edge-density of $k$-planar graphs is not completely settled.  For $k\le 4$, the maximum number of edges in an $n$-vertex $k$-planar graph is $(k+3)(n-2)$, which is tight for $k=1,2$ \cite{pach.toth:graphs}.  A near-tight bound of $11(n-2)/2$ is known for $k=3$ \cite{pach.radoicic.ea:improving} but for $k>4$, the current best upper bound is $4.108\sqrt{k}n$ \cite{pach.toth:graphs}.

% In the current paper we investigate the structure of $k$-planar graphs.  Since planarity is closed under taking minors, planar graphs can be characterised by a finite set of forbidden minors.  Indeed, Wagner's Theorem states that this set consists of the complete 5-vertex graph $K_5$ and the complete bipartite graph $K_{3,3}$.  Unfortunately, $k$-planarity is not closed under taking minors, ruling out this kind of clean structural result.  The purpose of the current paper is to extend a new structural result for planar graphs to $k$-planar graphs.

%The purpose of the current paper is to show that new structural results for planar graphs can be extended to $k$-planar graphs.  These results can be used to give a rough characterization of $k$-planar graphs of powers of bounded-degree planar graphs
% Rather, we present a rough structural characterization of $k$-planar graphs that, despite its roughness, provides several new results for $k$-planar and related graph families.  
%All of these results are based on layered $H$-partitions, which we now define.{\color{red}TODO: Help.}

%Thus, like genus-$g$ graphs,$k$-planar graphs are a generalization of planar graphs (which are genus-$0$ graphs).  Unlike genus-$g$ graphs, however, the family of $k$-planar graphs is not minor-closed.  A graph $G'$ obtained from a $k$-planar graph $G$ by edge deletions and edge contractions may or may not be $k$-planar. Indeed, \citet{dujmovic.eppstein.ea:structure} construct 1-planar graphs that contain arbitrarily large complete graph minors.

%\note{PM}{TODO: Write here about applications. non-repetitive colouring, $p$-centered colouring, $(g,k)$-planar graph, $(g,d)$-map graphs, $k$-NN graphs, David's rough characterization.}

%The remainder of the paper is organised as follows: In \cref{k-planar} we prove \cref{sec-k-planar}.  In \cref{sec-1-planar} we prove \cref{1-planar}.   In \cref{g-k-planar} we prove \cref{g-k-planar}. In \cref{consequences} we point out some consequences of these results for some of the other graph parameters mentioned above. \comment{Re-DO}

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Tools}
\label{Tools}

This section defines concepts from from the literature that will be important for our work.

In this paper, all graphs are finite and undirected. Unless specifically mentioned otherwise, all graphs are also simple. For any graph $G$ and any set $S$ (typically $S\subseteq V(G)$) we use $G[S]$ to denote the graph with vertex set $V(G)\cap S$ and edge set $\{uv\in E(G) : u,v\in S\}$.  We use $G-S$ as a shorthand for $G[V(G)\setminus S]$. We use $G'\subseteq G$ to denote subgraph containment; that is, $V(G')\subseteq V(G)$ and $E(G')\subseteq E(G)$.

We now formally define $k$-planar graphs.  An \emph{embedded graph} $G$ is a graph with $V(G)\subset\R^2$ in which each edge $vw\in E(G)$ is a closed curve\footnote{A closed curve in a surface $\Sigma$ is a continuous function $f:[0,1]\to \Sigma$. The points $f(0)$ and $f(1)$ are called the \emph{endpoints} of the curve.  When there is no danger of misunderstanding we treat a curve $f$ as the point set $\{f(t):0\le t\le 1\}$.} in $\R^2$ with endpoints $v$ and $w$ and not containing any vertex of $G$ in its interior.  A \emph{crossing} in an embedded graph $G$ is a triple $(p,vw,xy)$ with $p\in\R^2$, $vw,xy\in E(G)$ and such that $p\in (vw\cap xy)\setminus\{v,w,x,y\}$. An embedded graph $G$ is \emph{$k$-plane} if each edge of $G$ takes part in at most $k$ crossings.  A (not necessarily embedded) graph $G'$ is \emph{$k$-planar} if there exists $k$-plane graph $G$ isomorphic to $G'$.  Under these definitions, $0$-planar graphs are exactly planar graphs and $0$-plane graphs are exactly plane graphs.

We now define two concepts used in the statements of \cref{PlanarBasic,kPlanarBasic}: strong products and treewidth. The \emph{strong product} of graphs $A$ and $B$, denoted by $A\boxtimes B$, is the graph with vertex set $V(A)\times V(B)$, where distinct vertices $(v,x),(w,y)\in V(A)\times V(B)$ are adjacent if: 
\begin{compactitem}
\item  $v=w$ and $xy\in E(B)$, or 
\item  $x=y$ and $vw\in E(A)$, or  
\item  $vw\in E(A)$ and $xy\in E(B)$. 
\end{compactitem}
A \emph{tree-decomposition} $\mathcal{K}$ of a graph $G$ consists of a tree $K$ and a collection $\mathcal{K}=(B_x:x\in V(K))$ of subsets of $V(G)$ indexed by the nodes of $K$ such that:
\begin{compactenum}[(i)]
\item for every $vw\in E(G)$, there exists some $x\in V(K)$ with $v,w\in B_x$; and 
\item for every $v\in V(G)$, the induced subgraph $K[x] := K[\{y: x\in B_y\}]$ is connected.  
\end{compactenum}
The \emph{width} of the tree-decomposition $\mathcal{K}$ is $\max\{|B_x|:x\in V(K)\}-1$.  The \emph{treewidth} $\tw(G)$ of a graph $G$ is the minimum width of a tree-decomposition of $G$.  Treewidth is the standard measure of how similar a graph is to a tree. Indeed, a connected graph has treewidth 1 if and only if it is a tree. Treewidth is of fundamental importance in structural and algorithmic graph theory; see \citep{Reed03,HW17,Bodlaender-TCS98} for surveys.

While strong products enable an easy statement of \cref{PlanarBasic,kPlanarBasic}, to prove such theorems it is helpful to work with layerings and partitions, which we now introduce. A \emph{layering} of a graph $G$ is a sequence $\mathcal{L}=\langle V_0,V_1,\ldots\rangle$ such that $\{V_0,V_1,\ldots\}$ is a partition of $V(G)$ and for every edge $vw\in E(G)$, if $v\in V_i$ and $w\in V_j$ then $|j-i|\leq 1$.  For any partition $\mathcal{P}=\{S_1,\ldots,S_p\}$ of $V(G)$, a \emph{quotient graph} $H=G/\mathcal{P}$ has a $p$-element vertex set $V(H)=\{x_1,\ldots,x_p\}$ and $x_ix_j\in E(H)$ if and only if there exists an edge $vw\in E(G)$ such that $v\in S_i$ and $w\in S_j$. To highlight the importance of the quotient graph $H$, we call $\mathcal{P}$ an \emph{$H$-partition} and write this concisely as $\mathcal{P}=\{S_x : x\in V(H)\}$ so that each element of $\mathcal{P}$ is indexed by the vertex it creates in $H$.\footnote{An alternative definition of a layering of $G$ is an $H$-partition of $G$ where $H$ is a path.}  A layered $H$-partition $(\mathcal{L},\mathcal{P})$ of a graph $G$ consists of a layering $\mathcal{L}$ of $G$ and an $H$-partition $\mathcal{P}$ of $G$. The \emph{layered width} of $(\mathcal{L},\mathcal{P})$ is $\max\{|L\cap P|: L\in\mathcal{L},\, P\in\mathcal{P}\}$. These definitions relate to strong products as follows.

\begin{lem}[\citep{dujmovic.joret.ea:planar}] 
\label{PartitionProduct}
For every graph $H$, a graph $G$ has an $H$-partition of layered width at most $\ell$ if and only if $G$ is a subgraph of 
$H \boxtimes P \boxtimes K_\ell$ for some path $P$.
\end{lem}

\citet{dujmovic.joret.ea:planar} also showed it suffices to consider partitions of layered width 1.

\begin{lem}[\citep{dujmovic.joret.ea:planar}] 
\label{MakeWidth1}
If a graph $G$ has an $H$-partition of layered width $\ell$ with respect to layering $\langle V_0,V_1,\dots\rangle$, for some graph $H$ of treewidth at most $k$, then $G$ has an $H'$-partition of layered width 1 with respect to the same layering, for some graph $H'$ of treewidth at most $(k+1)\ell-1$.  That is, if $G\subseteq H\boxtimes P\boxtimes K_\ell$ for some graph $H$ of treewidth at most $k$  and for some path $P$, then $G\subseteq H' \boxtimes P$ for some graph $H'$ of treewidth at most $(k+1)\ell-1$.
\end{lem}

\citet{dujmovic.joret.ea:planar} proved the following result, which with \cref{PartitionProduct,MakeWidth1}, implies \cref{PlanarBasic}.

\begin{thm}[\citep{dujmovic.joret.ea:planar}]
\label{PlanarPartition}
Every planar graph has:
\begin{compactitem}
\item an $H$-partition of layered width $1$ for some planar graph $H$ of treewidth at most $8$, and
\item an $H$-partition of layered width $3$ for some planar graph $H$ of treewidth at most $3$.
\end{compactitem}
\end{thm}

%Layered $H$-partitions of small layered width in which $H$ has some additional property are a useful tool for a number of graph problems.  One such useful property is small \emph{treewidth}, which we now define.

% \citet{dujmovic.joret.ea:planar} show that, if each graph $G$ in some family $\mathcal{G}$ of graphs has a layered $H$-partition for which the layered width of the partition and the treewidth of $H$ are each upper-bounded by some constant $c_\mathcal{G}$, then each of the following quantities are upper bounded by a constant, for every $G\in\mathcal{G}$: queue number, track number, and layered treewidth. \citet{dujmovic.esperet.ea:planar} add non-repetitive chromatic number to this list.  \citet{dujmovic.joret.ea:planar} show that two common families of graphs have such $H$-partitions: planar graphs, graphs of bounded genus, and (more generally) apex-minor-free graphs.

% In this paper we show that another family of graphs admits layered $H$-partitions of bounded layered width in which $H$ has bounded treewidth:

\citet{dujmovic.joret.ea:planar} proved the following generalisation of \cref{PlanarBasic,PlanarPartition} for graphs embeddable on other surfaces.\footnote{The \textit{Euler genus} of the orientable surface with $h$ handles is $2h$. The \textit{Euler genus} of the non-orientable surface with $c$ cross-caps is $c$. The \textit{Euler genus} of a graph $G$ is the minimum integer $g$ such that $G$ embeds in a surface of Euler genus $g$. Of course, a graph is planar if and only if it has Euler genus 0; see \citep{mohar.thomassen:graphs} for more about graph embeddings in surfaces.}

\begin{thm}[\citep{dujmovic.joret.ea:planar}]
\label{SurfacePartition}
Every graph of Euler genus $g$ is a subgraph of:
\begin{compactenum}[(a)]
\item  $H \boxtimes P \boxtimes K_{\max\{2g,1\}}$ for some graph $H$ of treewidth at most $9$ and for some path $P$;
\item  $H \boxtimes P \boxtimes K_{\max\{2g,3\}}$ for some graph $H$ of treewidth at most $4$ and for some path $P$;
\item $(K_{2g} + H )  \boxtimes P$ for some graph $H$ of treewidth at most $8$  and some path $P$. 
\end{compactenum}
Equivalently, every graph of Euler genus $g$ has:
\begin{compactenum}[(a)]
\item an $H$-partition with layered width at most $\max\{2g,1\}$ such that $H$ has treewidth at most $9$;
\item an $H$-partition with layered width at most $\max\{2g,3\}$ such that $H$ has treewidth at most $4$; 
\item an $H$-partition with layered width at most $1$ such that $H$ has treewidth at most $2g+7$. 
\end{compactenum}
%\textcolor{red}{eliminate some of these results?}
\end{thm}

\citet{dujmovic.joret.ea:planar} generalised \cref{SurfacePartition} further to show that for every apex\footnote{A graph $M$ is a \textit{minor} of a graph $G$ if a graph isomorphic to $M$ can be obtained from a subgraph of $G$ by contracting edges. A class $\mathcal{G}$ of graphs is \emph{minor-closed} if for every graph $G\in\mathcal{G}$, every minor of $G$ is in $\mathcal{G}$. A minor-closed class is \emph{proper} if it is not the class of all graphs. For example, for fixed $g\geq 0$, the class of graphs with Euler genus at most $g$ is a proper minor-closed class. A graph $G$ is $t$-apex if it contains a set $A$ of at most $t$ vertices such that $G-A$ is planar. A 1-apex graph is \emph{apex}.  A minor-closed class $\mathcal{G}$ is apex-minor-free if some apex graph is not in $\mathcal{G}$.}
 graph $X$, there is a constant $c$ such that every $X$-minor-free graph is a subgraph of $H\boxtimes P$ for some graph $H$ of treewidth at most $c$. Apex-minor-free graphs are the largest minor-closed class for which such a theorem holds \citep{dujmovic.joret.ea:planar}. However, the family of $k$-planar graphs is not minor-closed.  A graph $G'$ obtained from a $k$-planar graph $G$ by edge deletions and edge contractions may or may not be $k$-planar. Indeed, \citet{dujmovic.eppstein.ea:structure} construct 1-planar graphs that contain arbitrarily large complete graph minors. Our results for $k$-planar graphs are the first of this type for a non-minor-closed class.

The following is the main result of this paper; it shows that $k$-planar graphs admit $H$-partitions with bounded layered width, in which $H$ has bounded treewidth. With \cref{PartitionProduct,MakeWidth1}, this proves \cref{kPlanarBasic}. % with $c_k\leq(4+o(1))k^5$.

\begin{thm}
\label{k-planar}
Every $k$-planar graph has a layered $H$-partition of layered width at most $24k^2 + 60k + 36$ in which $H$ has treewidth at most $\treewidth$.
\end{thm}

@@

\begin{thm}
\label{1-planar}
Every 1-planar graph has a layered $H$-partition of layered width at most 30 where $H$ is planar and has treewidth at most 3.
\end{thm}

%These results imply that $k$-planar graphs have all the graph parameters mentioned above upper-bounded by some constant $c_k$.
